---
title: "project 1"
author: "Andrew Tran, Shreya Dubey, Joseph Bannon"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
diamond <- read.csv("diamonds4.csv")
library(tidyverse)
library(ggplot2)
library(dplyr)
library(GGally)
library(MASS)
```

# Project 1: Report



## Section 1

```{r Data Transformation}

diamonds <- diamond %>%
              mutate(color_category = case_when(
                                        color %in% c("D", "E", "F") ~ "Colorless",
                                        color %in% c("G", "H", "I", "J") ~ "Near Colorless",
                                        color == "K" ~ "Faint",
                                        TRUE ~ "Other"),
                     clarity_category = case_when(
                                          clarity %in% c("I1", "I2", "I3") ~ "Included",
                                          clarity %in% c("SI1", "SI2") ~ "Slightly Included",
                                          clarity %in% c("VS1", "VS2") ~ "Very Slightly Included",
                                          clarity %in% c("VVS1", "VVS2") ~ "Very Very Slightly Included",
                                          clarity %in% c("IF") ~ "Internally Flawless",
                                          clarity %in% c("FL") ~ "Flawless",
                                          TRUE ~ "Other"))
              
```

```{r}
# Quick summary stats for diamond dataset variables
summary(diamonds)
```
Carat values range from 0.23 to 7, with a median value of 0.52 and mean value of 0.81.

Price values range from 322 to 355403, with a median value of 1463.5 and mean value of 7056.7.

The range of the numerical variables (carat and price) suggest the presence of outlier values that may skew values of central tendency. 

## Visualizing the Distribution of Numerical Variables

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() +
   geom_smooth(method = "lm", se=FALSE)
```

```{r Density Plots}
diamonds %>%
  ggplot(aes(price)) +
  geom_density()

diamonds %>%
  ggplot(aes(carat)) +
  geom_density()
```
```{r}

# Looking at distribution of smaller carat diamonds (below average)
diamonds %>% 
  filter(carat <= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

# Looking at distribution of larger carat diamonds (above average)
diamonds %>% 
  filter(carat >= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.10)
```

```{r}
# Looking at distribution of diamond price (below median)
diamonds %>% 
  filter(price <= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10)

# Looking at distribution of diamond price (above average)
diamonds %>% 
  filter(price >= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10000)
```
## Visualizing the Distribution of Categorical Variables

```{r}
diamonds %>%
ggplot(aes(x = fct_infreq(cut))) +
  geom_bar()

diamonds %>%
  group_by(cut) %>%
  tally() %>%
  arrange(desc(n))

diamonds %>%
ggplot(aes(x = fct_infreq(clarity))) +
  geom_bar()

diamonds %>%
  group_by(clarity, clarity_category) %>%
  tally() %>%
  arrange(desc(n))

diamonds %>%
ggplot(aes(x = fct_infreq(color))) +
  geom_bar()

diamonds %>%
  group_by(color, color_category) %>%
  tally() %>%
  arrange(desc(n))

```
## Visualizing Relationships

```{r Density Plot: Carat + Categoricals}

diamonds %>%
ggplot(aes(x = carat, color = cut)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = carat, color = clarity)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = carat, color = color)) +
  geom_density(linewidth = 0.75)
```

```{r Density Plots: Price + Categoricals}

diamonds %>%
ggplot(aes(x = price, color = cut)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = price, color = clarity)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = price, color = color)) +
  geom_density(linewidth = 0.75)
```

```{r}

diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) +
   geom_smooth(method = "lm", se=FALSE) +
  facet_wrap(~clarity_category, ncol = 3)

diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) +
   geom_smooth(method = "lm", se=FALSE) +
  facet_wrap(~color_category, ncol = 2)
```

```{r}
diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = color_category)) +
  geom_smooth(method = "lm", se=FALSE) +
  facet_wrap(~cut, ncol = 2)

diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = clarity_category)) +
  geom_smooth(method = "lm", se=FALSE)+
  facet_wrap(~cut, ncol = 2)
```

```{r}

# Looking at distribution of smaller carat diamonds (below average)
diamonds %>% 
  filter(carat <= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

# Looking at distribution of larger carat diamonds (above average)
diamonds %>% 
  filter(carat >= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.10)
```

```{r}
# Looking at distribution of diamond price (below median)
diamonds %>% 
  filter(price <= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10)

# Looking at distribution of diamond price (above average)
diamonds %>% 
  filter(price >= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10000)
```

```{r}
ggplot(diamonds, aes(x = fct_reorder(cut, price, median), y = price)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(color_category, price, median), y = price)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(clarity_category, price, median), y = price)) +
  geom_boxplot()
```

```{r}
ggplot(diamonds, aes(x = fct_reorder(cut, carat, median), y = carat)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(clarity_category, carat, median), y = carat)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(color_category, carat, median), y = carat)) +
  geom_boxplot()
```

## Looking at Covariation Between Two Categorical Variables

```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))

diamonds %>% 
  count(cut, clarity_category) %>%  
  ggplot(aes(x = cut, y = clarity_category)) +
  geom_tile(aes(fill = n))

diamonds %>% 
  count(color_category, clarity_category) %>%  
  ggplot(aes(x = color_category, y = clarity_category)) +
  geom_tile(aes(fill = n))
```

## SLR Model Diagnostics

```{r}
diamonds_slr <-lm(price~carat, data=diamonds)
par(mfrow = c(2, 2))
plot(diamonds_slr)

MASS::boxcox(diamonds_slr)
```

## Section 2


## Section 3


```{r}
diamond$cut <- as.factor(diamond$cut)
levels(diamond$cut)<-c( "Good","Very Good", "Ideal", "Astor Ideal") 
```

```{r}
diamond$clarity <- as.factor(diamond$clarity)
levels(diamond$clarity) <- c("SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF", "FL")
```

```{r}
diamond$color <- as.factor(diamond$color)
levels(diamond$color) <- rev(LETTERS[4:11]) 
```



