---
title: "project 1"
author: "Andrew Tran, Shreya Dubey, Joseph Bannon"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
diamond <- read.csv("diamonds4.csv")
library(tidyverse)
library(ggplot2)
library(dplyr)
library(GGally)
library(MASS)
```

# Project 1: Report


## Section 1

## Section 2

```{r Data Transformation}
diamonds <- diamonds %>%
              mutate(color_category = case_when(
                                        color %in% c("D", "E", "F") ~ "Colorless",
                                        color %in% c("G", "H", "I", "J") ~ "Near Colorless",
                                        color == "K" ~ "Faint",
                                        TRUE ~ "Other"),
                     clarity_category = case_when(
                                          clarity %in% c("I1", "I2", "I3") ~ "Included",
                                          clarity %in% c("SI1", "SI2") ~ "Slightly Included",
                                          clarity %in% c("VS1", "VS2") ~ "Very Slightly Included",
                                          clarity %in% c("VVS1", "VVS2") ~ "Very Very Slightly Included",
                                          clarity %in% c("IF") ~ "Internally Flawless",
                                          clarity %in% c("FL") ~ "Flawless",
                                          TRUE ~ "Other"))

head(diamonds)
              
```


```{r}
# Quick summary stats for diamond dataset variables
summary(diamonds)
```
Carat values range from 0.23 to 7, with a median value of 0.52 and mean value of 0.81.

Price values range from 322 to 355403, with a median value of 1463.5 and mean value of 7056.7.

The range of the numerical variables (carat and price) suggest the presence of outlier values that may skew values of central tendency. 

## Visualizing the Distribution of Numerical Variables

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point() +
   geom_smooth(method = "lm", se=FALSE)
```

```{r Density Plots}
diamonds %>%
  ggplot(aes(price)) +
  geom_density()

diamonds %>%
  ggplot(aes(carat)) +
  geom_density()
```
```{r}

# Looking at distribution of smaller carat diamonds (below average)
diamonds %>% 
  filter(carat <= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

# Looking at distribution of larger carat diamonds (above average)
diamonds %>% 
  filter(carat >= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.10)
```

```{r}
# Looking at distribution of diamond price (below median)
diamonds %>% 
  filter(price <= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10)

# Looking at distribution of diamond price (above average)
diamonds %>% 
  filter(price >= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10000)
```
## Visualizing the Distribution of Categorical Variables

```{r}
diamonds %>%
ggplot(aes(x = fct_infreq(cut))) +
  geom_bar()

diamonds %>%
  group_by(cut) %>%
  tally() %>%
  arrange(desc(n))

diamonds %>%
ggplot(aes(x = fct_infreq(clarity))) +
  geom_bar()

diamonds %>%
  group_by(clarity, clarity_category) %>%
  tally() %>%
  arrange(desc(n))

diamonds %>%
ggplot(aes(x = fct_infreq(color))) +
  geom_bar()

diamonds %>%
  group_by(color, color_category) %>%
  tally() %>%
  arrange(desc(n))

```
## Visualizing Relationships

```{r Density Plot: Carat + Categoricals}

diamonds %>%
ggplot(aes(x = carat, color = cut)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = carat, color = clarity)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = carat, color = color)) +
  geom_density(linewidth = 0.75)
```

```{r Density Plots: Price + Categoricals}

diamonds %>%
ggplot(aes(x = price, color = cut)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = price, color = clarity_category)) +
  geom_density(linewidth = 0.75)

diamonds %>%
ggplot(aes(x = price, color = color_category)) +
  geom_density(linewidth = 0.75)
```

```{r}

diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) +
   geom_smooth(method = "lm", se=FALSE) +
  facet_wrap(~clarity_category, ncol = 3)

diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) +
   geom_smooth(method = "lm", se=FALSE) +
  facet_wrap(~color_category, ncol = 2)
```

```{r}
diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = color_category)) +
  geom_smooth(method = "lm", se=FALSE) +
  facet_wrap(~cut, ncol = 2)

diamonds %>%
ggplot(aes(x = carat, y = price)) +
  geom_point(aes(color = clarity_category)) +
  geom_smooth(method = "lm", se=FALSE)+
  facet_wrap(~cut, ncol = 2)
```

```{r}

# Looking at distribution of smaller carat diamonds (below median)
diamonds %>% 
  filter(carat <= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.01)

# Looking at distribution of larger carat diamonds (above median)
diamonds %>% 
  filter(carat >= median(carat)) %>%
  ggplot(aes(x = carat)) +
  geom_histogram(binwidth = 0.10)
```

```{r}
# Looking at distribution of diamond price (below median)
diamonds %>% 
  filter(price <= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10)

# Looking at distribution of diamond price (above average)
diamonds %>% 
  filter(price >= median(price)) %>%
  ggplot(aes(x = price)) +
  geom_histogram(binwidth = 10000)
```

```{r}
ggplot(diamonds, aes(x = fct_reorder(cut, price, median), y = price)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(color_category, price, median), y = price)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(clarity_category, price, median), y = price)) +
  geom_boxplot()
```

```{r}
ggplot(diamonds, aes(x = fct_reorder(cut, carat, median), y = carat)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(clarity_category, carat, median), y = carat)) +
  geom_boxplot()

ggplot(diamonds, aes(x = fct_reorder(color_category, carat, median), y = carat)) +
  geom_boxplot()
```

## Looking at Covariation Between Two Categorical Variables

```{r}
diamonds %>% 
  count(color, cut) %>%  
  ggplot(aes(x = color, y = cut)) +
  geom_tile(aes(fill = n))

diamonds %>% 
  count(cut, clarity_category) %>%  
  ggplot(aes(x = cut, y = clarity_category)) +
  geom_tile(aes(fill = n))

diamonds %>% 
  count(color_category, clarity_category) %>%  
  ggplot(aes(x = color_category, y = clarity_category)) +
  geom_tile(aes(fill = n))
```

## SLR Model Diagnostics

```{r}
diamonds_slr <-lm(price~carat, data=diamonds)
par(mfrow = c(2, 2))
plot(diamonds_slr)

MASS::boxcox(diamonds_slr, lambda = seq(-0.5, 0.5, 1/10))
```
## Variable Transformation and Assumption Testing

```{r}

# Log transformation of y (price)
diamonds <- diamonds %>%
              mutate(ystar = log(price))

diamonds_slr_ystar <-lm(ystar~carat, data=diamonds)
par(mfrow = c(2, 2))
plot(diamonds_slr_ystar)

# log transformation of x (carat)

diamonds <- diamonds %>%
              mutate(xstar = log(carat))
```
Single Linear Regression for carat and price

```{r}
diamonds_slr_xystar <-lm(ystar~xstar, data=diamonds)
par(mfrow = c(2, 2))
plot(diamonds_slr_xystar)

summary(diamonds_slr_xystar)
```

## SLR Analysis

### T Test

H0: There is no linear relationship between carat and price.
H1: There is a linear relationship between carat and price.

t value for carat(log transformed) is 159.8 with a p-value of < 2.2e-16. Null hypothesis is rejected and we can conclude that there is a statistically significant linear relationship between the carat and price.

### ANOVA Testing

```{r ANOVA }
anova(diamonds_slr_xystar)
```
#### Coefficient of Determination

```{r}
anova.tab<-anova(diamonds_slr_xystar)
##SST not provided, so we add up SSR and SSres
SST<-sum(anova.tab$"Sum Sq")
##R2
anova.tab$"Sum Sq"[1]/SST
```
Proportion of variance in price that can be explained carat is 0.9546863

#### Correlation between Price and Carat

```{r}
cor(diamonds$ystar, diamonds$xstar)
```

This value indicates a very strong and positive linear association between price and carat.

### Confidence Interval for Regression Coefficient

```{r}
## 95% CIs for all regression coefficients
confint(diamonds_slr_xystar,level = 0.95)
```

The 95% CI for carat is (1.920152, 1.967888). We have 95% confidence that for each unit increase in diamond carat, the predicted price of diamond increases 8.52 to 8.54 (???)

#### Visualization of CI for mean response given x and PI of response given x

```{r}
##regular scatterplot
##with regression line overlaid, and bounds of CI for mean y

diamonds %>%
  ggplot(aes(x=xstar, y = ystar)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(x = "Carat (log)",
       y = "Price (log)",
       title = "Scatterplot of Carat (log) against Price (log)")

##find PIs for each observation
preds <- predict(diamonds_slr_xystar, interval="prediction")


diamonds <- data.frame(diamonds,preds)

##overlay PIs via geom_line()

diamonds %>%
  ggplot(aes(x=xstar, y = ystar)) +
  geom_point() +
  geom_line(aes(y=lwr), color = "red", linetype = "dashed")+
  geom_line(aes(y=upr), color = "red", linetype = "dashed")+
  geom_smooth(method = lm) +
  labs(x = "Carat (log)",
       y = "Price (log)",
       title = "Scatterplot of Carat (log) against Price (log)")
```

## MLR Analysis

```{r}
##convert categorical predictor to factor with ordered levels

diamonds <- diamonds %>%
              mutate(color = factor(color, levels = c("K", "J", "I", "H", "G", "F", "E", "D")),
                     clarity = factor(clarity),
                     color_category = factor(color_category, levels = c("Faint", "Near Colorless", "Colorless")),
                     clarity_category = factor(clarity_category,
                                               levels = c("Included", 
                                                         "Slightly Included", 
                                                         "Very Slightly Included", 
                                                         "Very Very Slightly Included", 
                                                         "Internally Flawless", 
                                                         "Flawless")),
                     cut = factor(cut, levels = c("Good", "Very Good", "Ideal", "Astor Ideal")))
```

### Analysis of Diamond Color on Price and Carat

```{r MLR Color}
## scatterplot with separate regression lines
diamonds %>%
  ggplot(aes(x=xstar, y=ystar, color=color_category))+
  geom_point()+
  geom_smooth(method=lm, se=FALSE)+
  labs(title="Scatterplot of Diamond Carat (log transformed) and Price (log transformed) Relationship, by Color Category")

# Model with additive effect of color category
diamonds_carat_price_color_mlr <- lm(ystar~xstar+color_category, data=diamonds)
summary(diamonds_carat_price_color_mlr)
```

#### ANOVA Linear F-test for Diamond Color

```{r MLR Color: ANOVA F-Test}
anova(diamonds_slr_xystar, lm(ystar~xstar+color_category, data=diamonds))
```
H0: The coefficients for levels of color are zero
H1: At least one coefficient for color is not zero


The p-value of for the model with color category being < 0.05 shows a statistically significant association between price and color for diamonds of the same carat size.

#### Tukey's Procedure for Color Categories

```{r MLR Color: Tukey}
color_tukey <- multcomp::glht(lm(ystar~xstar+color_category, data=diamonds), linfct = mcp(color_category = "Tukey"))
summary(color_tukey)
plot(color_tukey)
```
H0: There is no difference in average price between the color classifications of the same carat size
H1: There is a significant difference in average price between color classifications of the same carat size

p-values for color classification pairings are <0.05, which allows us to conclude that for each pairing, there is a statistically significant difference in average price between color classifications for the same carat diamond size. 

### Analysis of Diamond Clarity on Price and Carat

```{r}
##scatterplot with separate regression lines
diamonds %>%
  ggplot(aes(x=xstar, y=ystar, color=clarity_category))+
  geom_point()+
  geom_smooth(method=lm, se=FALSE)+
  labs(title="Scatterplot of Diamond Carat (log transformed) and Price (log transformed) Relationship, by Clarity Category")
```

```{r MLR Clarity}
# Model with additive effect of clarity category
diamonds_carat_price_clarity_mlr <- lm(ystar~xstar+clarity_category, data=diamonds)
summary(diamonds_carat_price_clarity_mlr)
```

#### ANOVA Linear F-test for Diamond Clarity

```{r MLR Clarity: ANOVA F-Test}
anova(diamonds_slr_xystar, lm(ystar~xstar+clarity_category, data=diamonds))
```
H0: The coefficients for all levels of clarity are zero
H1: At least one coefficient for clarity is not zero

The p-value of for the model with clarity category being < 0.05 shows a statistically significant association between price and clarity for diamonds of the same carat size.

```{r MLR Clarity: Tukey}
clarity_tukey <- multcomp::glht(lm(ystar~xstar+clarity_category, data=diamonds), linfct = mcp(clarity_category = "Tukey"))
summary(clarity_tukey)
plot(clarity_tukey)
```
H0: There is no difference in average price between the clarity classifications of the same carat size
H1: There is a significant difference in average price between clarity classifications of the same carat size

p-values for clarity classification pairings are <0.05, which allows us to conclude that for each pairing, there is a statistically significant difference in average price between clarity classifications for the same carat diamond size. 

### Analysis of Diamond Cut on Price and Carat

```{r Scatterplot: Price, Carat, Cut}
##scatterplot with separate regression lines
diamonds %>%
  ggplot(aes(x=xstar, y=ystar, color=cut))+
  geom_point()+
  geom_smooth(method=lm, se=FALSE)+
  labs(title="Scatterplot of Diamond Carat (log transformed) and Price (log transformed) Relationship, by Cut Category")
```


```{r MLR Cut}
# Model with additive effect of color category
diamonds_carat_price_cut_mlr <- lm(ystar~xstar+cut, data=diamonds)

summary(diamonds_carat_price_cut_mlr)
```

#### ANOVA Linear F-test for Diamond Cut

```{r MLR Cut: ANOVA F-Test}
anova(lm(ystar~xstar, data=diamonds), lm(ystar~xstar*cut, data=diamonds))
```
H0: The coefficients for levels of cut are zero
H1: At least one coefficient for cut is not zero

With a p-value < 0.05, the null hypothesis can be rejected and conclude that at least one coefficient for diamond cut is not zero and can improve the model's ability to predict price. We can conclude that diamond cut, as a whole, has a statistically significant association with price, even after accounting for carat size.

#### Tukey's Procedure for Cut Categories

```{r MLR Cut: Tukey}
cut_tukey <- multcomp::glht(diamonds_carat_price_cut_mlr, linfct = mcp(cut = "Tukey"))
summary(cut_tukey)
plot(cut_tukey)
```
H0: There is no difference in average price between the cuts of the same carat size
H1: There is a significant difference in average price between cuts of the same carat size

Tukey's procedure for evaluating diamond cuts on price for diamonds of the same carat size showed significant differences in Ideal & Good, Astor Ideal & Good, Ideal & Very Good, and Astor Ideal & Very Good pairs. Very Good & Good and Astor Ideal and & Ideal pairs did not have significant differences in price for diamonds of the same carat size. 





## Section 3

### Intial Regression Assumption

To verify the initial regression assumptions are met, we will examine a scatter plot and residual plot.

```{r}
ggplot(diamond, aes(x = carat, y = price)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Price (dollars) vs. Carat",
       x = "Carat",
       y = "Price (dollars)")
```
In this case, the predictor variable is Carat and the response is Price (dollars). The assumption that mean error is 0 (i.e.linear data) seems not to hold true as the data does not look very linear. For example the observations of price between 0-.5 and 5-7 of carat is all above the regression line while the rest of the observations of price are bellow regression line. This means the SLR estimations of price will systemically under predict or over predict price in the respective regions.

Additionally, the assumption of constant variance also does not look to hold true as the variance seems to be higher for higher values of carat. This means statistical inference test for hypothesis tests and confidence intervals can not be used until variance is stabilized. 



```{r}
result<-lm(price ~ carat, data=diamond)
par(mfrow = c(2, 2))
plot(result)
```

Looking at the "Residual vs Fitted" plot, the vertical spread seems to be higher for higher fitted values indicating variance is not constant. Furthermore, the plot seems to indicate a non-linear relationship which indicates mean error is 0. Therefore, the first transformation we will try is on the response variable (Price). This will stabilize the variance of the response so we can assess assumption of mean error is 0 more accurately.

The "Scale-Location" plot also contributes additional evidence against the assumptions that variance is not constant as the plot shows curved fitted line and increasing variances moving left to right on the graph.

The "Q-Q Residuals" plot indicates that that errors are not normally distributed because the values do not fall on a roughly 45 degree line. However, since this assumption is the least important, it is ignored until the other assumptions are met. 


```{r}
acf(result$residuals, main="ACF Plot of Residuals with Price")
```

Given the "ACF Plot" the assumptions that errors are independent seems to hold true as lag is mostly under the critical value (expect for value 3).


### Transformation of variables

```{r}
MASS::boxcox(result, lambda = seq(.2, 0.4, 1/10)) 
```

The box-cox plot maximizes log-likelihood function against lambda to determine which transformation will likely stabilize the variance of response. Using a value inside the outer dashed lines (95% CI) is the most likely transformation of the data to stabilize variance. I will choose .3 since it is inside the 95% CI and closest to the predicted optimal value for lambda. Log transformation would have been preferred because log transformation preserve the meaning of the coefficients. However, since $\lambda = 0$ is not in the 95% CI, this is not a viable transformation.


```{r}
ystar<-(diamond$price)^(.3)
Data<-data.frame(diamond,ystar)
```

```{r}
ggplot(Data, aes(x = carat, y = ystar)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Price (transformed) vs. Carat",
       x = "Carat",
       y = "Price (transformed)")
```



```{r}
result.ystar<-lm(ystar~carat, data=Data)
par(mfrow = c(2, 2))
plot(result.ystar)
```


Looking at the "Residual vs Fitted" and "Scale-Location" plot, the variance of data around the fitted line is relatively constant moving left to right on the plot. This indicates the transformation on the response succeeded in stabilizing the variance so the constant variance of errors is met. 

However, the "Residual vs Fitted" plot still seems to indicate a non linear relationship because the red line is not straight so a transformation on the predictor is the next step.

The un-transformed predictor plot above looks roughly like a square-root plot so we will attempt that transformation and assess residuals.


```{r}
Data$xstar<- (Data$carat)^(1/2)
```

```{r}
ggplot(Data, aes(x = xstar, y = ystar)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Price (transformed) vs. Carat (transformed)",
       x = "Carat (transformed)",
       y = "Price (transformed)")
```

```{r}
result.xystar<-lm(ystar~xstar, data=Data)
par(mfrow = c(2, 2))
plot(result.xystar)
```

The square root transformation on the predictor seems to have improved the "Residual vs Fit" line so that it is straight and horizontal indicating that assumption  mean error is 0 is met.

The assumption of constant variance seems to be mostly met as the variance of the data in the "Residual vs Fit" graph seems to be primarily constant over the values of the response variable. Also the "Scale-Location" line has a slight slope, but this still supports the conclusion assumption of constant variance is met.

The "Q-Q Residuals" plot indicates that that errors are more normally distributed after the transformation by comparing the plot before vs after the transformation, the values are closer to the 45 degree line with some variation in each end. However, since this assumption is the least important, we can make the assumption that it is met. 

```{r}
acf(result.xystar$residuals, main="ACF Plot of Residuals with Price")
```

Given the "ACF Plot", the assumptions that errors are independent is met as all lag values are under the critical value.


Now that the four regression assumptions are met, we can fit a Simple Linear Regression model to our transformed data.


### Simple Linear Regression

```{r}
result_final<-lm(ystar~xstar, data=Data)
summary(result_final)
```
Simple Linear Regression equation (where :

$\hat{Price}^{.3} = -2.65220 + 15.91060 * Carat^{1/2}$

Unfortunately the value of the parameters can not be interpreted because the transformation were not logarithmic. However, contextually the positive slope indicates that as the carat increase, the price increase. This relationship is non linear as seen by the transformations applied in the previous section. From the transformation applied to predictor it seems that as carat increase, price increases roughly quadratically (i.e. the square root transformation to linearize the data).


After viewing the results of the hypothesis test, we can conclude that the results for (transformed) slope and intercept are statistically significant because they have p value much less than .05 (p<2e-16). 

The confidence intervals for slope are as follows:

```{r}
confint(result_final,level = 0.95)[2,]

```

The 95% confidence interval for slope is (15.69645, 16.12474) confirming the hypothesis test that the (transformed) relationship between Price and Carat is linear because the Confidence Interval does not include 0. 




