---
title: "project 1"
author: "Andrew Tran, Shreya Dubey, Joseph Bannon"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
diamond <- read.csv("diamonds4.csv")
library(tidyverse)
library(ggplot2)
library(dplyr)
library(GGally)
```

# Project 1: Report


## Section 1


## Section 2


## Section 3

### Intial Regression Assumption

To verify the initial regression assumptions are met, we will examine a scatter plot and residual plot.

```{r}
ggplot(diamond, aes(x = carat, y = price)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Price (dollars) vs. Carat",
       x = "Carat",
       y = "Price (dollars)")
```
In this case, the predictor variable is Carat and the response is Price (dollars). The assumption that mean error is 0 (i.e.linear data) seems not to hold true as the data does not look very linear. For example the observations of price between 0-.5 and 5-7 of carat is all above the regression line while the rest of the observations of price are bellow regression line. This means the SLR estimations of price will systemically under predict or over predict price in the respective regions.

Additionally, the assumption of constant variance also does not look to hold true as the variance seems to be higher for higher values of carat. This means statistical inference test for hypothesis tests and confidence intervals can not be used until variance is stabilized. 



```{r}
result<-lm(price ~ carat, data=diamond)
par(mfrow = c(2, 2))
plot(result)
```

Looking at the "Residual vs Fitted" plot, the vertical spread seems to be higher for higher fitted values indicating variance is not constant. Furthermore, the plot seems to indicate a non-linear relationship which indicates mean error is 0. Therefore, the first transformation we will try is on the response variable (Price). This will stabilize the variance of the response so we can assess assumption of mean error is 0 more accurately.

The "Scale-Location" plot also contributes additional evidence against the assumptions that variance is not constant as the plot shows curved fitted line and increasing variances moving left to right on the graph.

The "Q-Q Residuals" plot indicates that that errors are not normally distributed because the values do not fall on a roughly 45 degree line. However, since this assumption is the least important, it is ignored until the other assumptions are met. 


```{r}
acf(result$residuals, main="ACF Plot of Residuals with Price")
```

Given the "ACF Plot" the assumptions that errors are independent seems to hold true as lag is mostly under the critical value (expect for value 3).


### Transformation of variables

```{r}
MASS::boxcox(result, lambda = seq(.2, 0.4, 1/10)) 
```

The box-cox plot maximizes log-likelihood function against lambda to determine which transformation will likely stabilize the variance of response. Using a value inside the outer dashed lines (95% CI) is the most likely transformation of the data to stabilize variance. I will choose .3 since it is inside the 95% CI and closest to the predicted optimal value for lambda. Log transformation would have been preferred because log transformation preserve the meaning of the coefficients. However, since $\lambda = 0$ is not in the 95% CI, this is not a viable transformation.


```{r}
ystar<-(diamond$price)^(.3)
Data<-data.frame(diamond,ystar)
```

```{r}
ggplot(Data, aes(x = carat, y = ystar)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Price (transformed) vs. Carat",
       x = "Carat",
       y = "Price (transformed)")
```



```{r}
result.ystar<-lm(ystar~carat, data=Data)
par(mfrow = c(2, 2))
plot(result.ystar)
```


Looking at the "Residual vs Fitted" and "Scale-Location" plot, the variance of data around the fitted line is relatively constant moving left to right on the plot. This indicates the transformation on the response succeeded in stabilizing the variance so the constant variance of errors is met. 

However, the "Residual vs Fitted" plot still seems to indicate a non linear relationship because the red line is not straight so a transformation on the predictor is the next step.

The un-transformed predictor plot above looks roughly like a square-root plot so we will attempt that transformation and assess residuals.


```{r}
Data$xstar<- (Data$carat)^(1/2)
```

```{r}
ggplot(Data, aes(x = xstar, y = ystar)) +
  geom_point() +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Price (transformed) vs. Carat (transformed)",
       x = "Carat (transformed)",
       y = "Price (transformed)")
```

```{r}
result.xystar<-lm(ystar~xstar, data=Data)
par(mfrow = c(2, 2))
plot(result.xystar)
```

The square root transformation on the predictor seems to have improved the "Residual vs Fit" line so that it is straight and horizontal indicating that assumption  mean error is 0 is met.

The assumption of constant variance seems to be mostly met as the variance of the data in the "Residual vs Fit" graph seems to be primarily constant over the values of the response variable. Also the "Scale-Location" line has a slight slope, but this still supports the conclusion assumption of constant variance is met.

The "Q-Q Residuals" plot indicates that that errors are more normally distributed after the transformation by comparing the plot before vs after the transformation, the values are closer to the 45 degree line with some variation in each end. However, since this assumption is the least important, we can make the assumption that it is met. 

```{r}
acf(result.xystar$residuals, main="ACF Plot of Residuals with Price")
```

Given the "ACF Plot", the assumptions that errors are independent is met as all lag values are under the critical value.


Now that the four regression assumptions are met, we can fit a Simple Linear Regression model to our transformed data.


### Simple Linear Regression

```{r}
result_final<-lm(ystar~xstar, data=Data)
summary(result_final)
```
Simple Linear Regression equation (where :

$\hat{Price}^{.3} = -2.65220 + 15.91060 * Carat^{1/2}$

Unfortunately the value of the parameters can not be interpreted because the transformation were not logarithmic. However, contextually the positive slope indicates that as the carat increase, the price increase. This relationship is non linear as seen by the transformations applied in the previous section. From the transformation applied to predictor it seems that as carat increase, price increases roughly quadratically (i.e. the square root transformation to linearize the data).


After viewing the results of the hypothesis test, we can conclude that the results for (transformed) slope and intercept are statistically significant because they have p value much less than .05 (p<2e-16). 

The confidence intervals for slope are as follows:

```{r}
confint(result_final,level = 0.95)[2,]

```

The 95% confidence interval for slope is (15.69645, 16.12474) confirming the hypothesis test that the (transformed) relationship between Price and Carat is linear because the Confidence Interval does not include 0. 




